%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% MSc Artificial Intelligence Thesis — University of Amsterdam
% Student: Mohammad Hafeez Khan
% Title: Semantic Scene Completion with Latent Modeling of Scene-Level
%        3D Gaussians
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[license]{mscaithesis}
\raggedbottom
\makeatletter
\@openrightfalse
\makeatother

\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{url}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}

% Shorthands
\newcommand{\GS}{\textsc{3dgs}}
\newcommand{\cL}{\mathcal{L}}
\newcommand{\cG}{\mathcal{G}}
\newcommand{\bR}{\mathbb{R}}
\newcommand{\bS}{\mathbb{S}}
\newcommand{\vmu}{\boldsymbol{\mu}}
\newcommand{\vsigma}{\boldsymbol{\Sigma}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\repo}{\textcolor{blue}{\url{https://github.com/Hafeezmhk7/Semantic-Scene-Completion.git}}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% METADATA
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\title{Semantic Scene Completion with Latent\\
       Modeling of Scene-Level 3D Gaussians}

\date{2026/06/19}  % TODO: update to actual submission date

\author{Mohammad Hafeez \textsc{Khan}}
\studentnumber{15549089}
\period{January 2025 -- May 2026}
\credits{36 ECTS}
\examiner{Dr. Shaodi You}   % Fill from Datanose
\dailysupervisor{~Yue Li}
\uvasupervisor{Prof.~Martin R.~Oswald}
\addsupervisor{}

\begin{document}
\maketitle

\frontmatter

    \chapter*{Abstract}
        \begin{adjustwidth}{4cm}{4cm}
            Photorealistic three-dimensional scene generation has, in the past two years, become both more attractive and more tractable thanks to 3D Gaussian Splatting (3DGS). A natural way to build a generative model on top of this representation is the standard two-stage recipe: first compress scenes into a structured latent with a variational autoencoder, then learn a flow-matching or diffusion model in that latent space. Existing work has shown this is possible at the level of small object-centric collections of a few hundred scenes. Scaling the same recipe up to several thousand pre-fitted indoor scenes is what this thesis sets out to study.

I begin from the architecture of Can3Tok \citep{gao2025can3tok} and use the SceneSplat-7K dataset \citep{li2025scenesplat} for training. Through a controlled ablation of seven Stage~1 variants, I show that two simple choices control whether the autoencoder converges or memorises at scale: (1) imposing a fixed-frame Hilbert ordering on the input Gaussians, which makes the slot-aligned reconstruction target identifiable, and (2) replacing the global flat decoder bottleneck with a windowed local encoder paired with a shared per-token MLP. Together they bring validation reconstruction down from a non-convergent regime to a level where qualitative reconstruction of held-out scenes becomes meaningful. A small-scale (300-scene) exploration of latent disentanglement and per-Gaussian contrastive supervision motivates a schema-aware Stage~2 design: a hierarchical flow-matching generator with a layout transformer and a geometry transformer for the disentangled variants, and a single scene-level transformer for the flat variants. Stage~2 training and the corresponding scene completion experiments are in progress at the time of writing.
        \end{adjustwidth}

    \tableofcontents
    \listoffigures
    \listoftables
    \setlength{\parskip}{5pt}
    \include{sections/0-symbols}

\twocolumn
\mainmatter
    \include{sections/1-introduction}
    \include{sections/2-related-work}
    \include{sections/3-methodology}
    \include{sections/4-results}
    \include{sections/5-discussion}
    \include{sections/6-conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{bibliographies/references}

\appendix
    \include{sections/7-appendix}

\end{document}


























