\chapter{Related Work}
\label{sec:related_work}

This thesis sits at the intersection of four lines of research: explicit 3D scene representation, latent generative modeling of 3D content, learning on unordered sets through serialization, and structured or semantically supervised representation learning. We review each in turn and, following the recommended structure, close each block by stating the gap it leaves for our work. The overarching gap is stated first and revisited at the end: there exists a scene-level 3DGS autoencoder that generalizes on a few hundred scenes, but no account of what is required to scale it, to make its latent semantically structured, or to make its reconstruction target identifiable.

% ---------------------------------------------------------------------
\section{Explicit 3D Scene Representation}
% ---------------------------------------------------------------------

Neural radiance fields introduced a continuous, implicit representation of a scene as a coordinate network queried along camera rays \citep{mildenhall2020nerf}. The implicitness, while elegant, is awkward for generative modeling: there is no explicit object to predict, only the weights of a per-scene network. 3D Gaussian Splatting replaced the implicit field with an explicit collection of anisotropic Gaussian primitives that are rasterized directly, achieving comparable or better quality with real-time rendering \citep{kerbl2023gaussian}. Each primitive carries a position $\mu \in \mathbb{R}^3$, an opacity $\alpha \in (0,1)$, a colour, and an anisotropic covariance $\Sigma$ factorized as a rotation (stored as a quaternion $q$) and a per-axis scale $s$, so that $\Sigma = R(q)\,\mathrm{diag}(s^2)\,R(q)^\top$.

The explicitness of 3DGS is what makes it a candidate substrate for generation, but it comes with three properties that complicate learning on it. First, a scene is a \emph{set} of primitives with no canonical order. Second, the fit is heavily over-parameterized, so the same scene can be represented by many different primitive configurations. Scaffold-GS \citep{lu2024scaffold} makes this redundancy explicit and partly tames it by attaching neural Gaussians to a sparse set of anchor points and predicting their attributes locally, which both reduces redundancy and imposes a coarse spatial organization; we borrow its anchor-plus-local-offset idea in our decoder. Third, the number of primitives is large, typically hundreds of thousands per scene. These properties (no order, redundancy, scale) are exactly what an autoencoder for 3DGS must absorb. The gap: 3DGS gives an explicit but unordered, redundant, and large representation, and a generative pipeline needs a compact and organized latent in its place.

% ---------------------------------------------------------------------
\section{Latent Generative Modeling of 3D Content}
% ---------------------------------------------------------------------

The dominant recipe for high-dimensional generation is two-stage latent modeling: an autoencoder compresses the signal, and a diffusion or flow-matching model is trained in the latent space \citep{rombach2022ldm}. The second-stage generator has matured from denoising diffusion \citep{ho2020ddpm} to flow matching \citep{lipman2023flowmatching} and rectified flow \citep{liu2023rectified}, which learn a continuous transport from a simple prior to the data and are typically realized with a diffusion transformer backbone \citep{peebles2023dit}. Our Stage 2 design adopts flow matching with a transformer backbone for these reasons.

For 3D \emph{objects}, this recipe is well established. LION trains a hierarchical latent point-cloud diffusion model \citep{zeng2022lion}. Michelangelo encodes a shape into a set of latent tokens with a Perceiver-style cross-attention encoder and learned query tokens, aligns that latent to image and text, and decodes an occupancy field with a transformer \citep{zhao2023michelangelo}; its shape-latent variational autoencoder is the architecture our encoder is built on. GaussianCube is the closest object-level analogue to our setting: it fits a fixed number of Gaussians per object and then transports them onto a regular voxel grid by optimal transport, so that a standard convolutional diffusion model can operate on a structured tensor \citep{zhang2024gaussiancube}. The voxel-grid structuring is powerful for bounded objects but does not transfer to unbounded scenes, where surfaces occupy a thin subset of a large volume and a dense grid is mostly empty; we adopt the spirit of its structuring (position as a structural reference plus a small offset) without the dense grid.

The common thread is that every one of these methods works in a bounded, canonical coordinate frame on object-scale data. Whole scenes are unbounded and scale-inconsistent, and feedforward scene-level 3D generation has barely been attempted \citep{gao2025can3tok}. The gap: the latent-generative recipe is proven at the object level but its prerequisite, a scene-level autoencoder that yields a well-behaved latent, is missing.

% ---------------------------------------------------------------------
\section{Learning on Unordered Sets and Serialization}
% ---------------------------------------------------------------------

Permutation invariance is the defining constraint of learning on point sets. PointNet established symmetric aggregation as the route to a permutation-invariant encoder \citep{qi2017pointnet}. The Perceiver and Perceiver IO generalized this with a fixed set of latent query tokens that attend to a large, arbitrarily-ordered input by cross-attention, decoupling the latent size from the input size \citep{jaegle2021perceiver,jaegle2022perceiverio}. Our encoder follows this pattern: a small set of learned queries cross-attends to the Gaussian set and yields a fixed-size latent regardless of how many primitives the scene contains.

Permutation invariance solves the \emph{encoder} but creates a problem for the \emph{decoder} objective. If the target is a set, a loss that compares output slot $i$ to target slot $i$ depends on an arbitrary correspondence and is, in general, not a learnable function. One response is permutation-invariant matching: Chamfer or earth-mover distances on point clouds, and the bipartite-matching loss of DETR, which assigns predictions to targets by optimal assignment before scoring \citep{carion2020detr}, computed efficiently with entropic optimal transport \citep{cuturi2013sinkhorn}. A second response is to impose a canonical order. Point Transformer V3 serializes a point cloud along a space-filling curve so that windowed attention becomes possible, and it finds that the Hilbert curve preserves locality better than the Z-order (Morton) curve, because consecutive positions on the Hilbert curve are always spatially adjacent \citep{wu2024ptv3,skilling2004hilbert}. Both responses are central to this thesis: we use a fixed-frame Hilbert serialization to make a slot-aligned loss learnable, and a Sinkhorn set-matching loss to remove the residual within-region assignment gauge. The gap: serialization and set matching are established tools for point clouds, but it is unclear how they interact, at the scale of thousands of scenes, with the additional gauge freedoms of per-primitive 3DGS parameters (orientation and within-region placement), or how the choice of how many primitives to sample affects them.

% ---------------------------------------------------------------------
\section{Structured and Semantically Supervised Representations}
% ---------------------------------------------------------------------

A latent that is merely reconstructive need not be organized. Two bodies of work aim to add structure. Contrastive learning shapes a representation by pulling together related instances and pushing apart unrelated ones, from contrastive predictive coding \citep{oord2018cpc} to SimCLR \citep{chen2020simclr}; we adapt the InfoNCE objective to the per-primitive level, pulling together Gaussians that share a semantic label. Disentanglement seeks latent axes that correspond to distinct factors of variation, classically through the $\beta$-VAE \citep{higgins2017betavae}, though later analysis showed that fully unsupervised disentanglement is generally unidentifiable without inductive biases or weak supervision \citep{locatello2019challenging}; we therefore pursue a weak, structural form of layout-versus-geometry separation rather than strict factorization.

The supervision needed to make a 3DGS latent semantic has only recently become available. SceneSplat introduced SceneSplat-7K, a dataset of $7{,}916$ indoor 3DGS scenes with per-primitive language and category labels, and a native 3DGS encoder for open-vocabulary scene understanding \citep{li2025scenesplat}; the follow-up SceneSplat++ extends the benchmark and data \citep{ma2025scenesplatpp}. These provide exactly the per-Gaussian labels our semantic objectives consume. The gap: dense per-primitive semantics exist, but they have been used for discriminative scene understanding, not to structure a \emph{generative} scene latent.

% ---------------------------------------------------------------------
\section{The Immediate Predecessor and the Remaining Gap}
% ---------------------------------------------------------------------

Can3Tok is the first variational autoencoder shown to encode scene-level 3DGS into a low-dimensional latent that generalizes to novel scenes \citep{gao2025can3tok}. It canonicalizes scenes with a normalization pipeline, tokenizes the Gaussian set with a Michelangelo-style cross-attention encoder, and decodes the primitives from the latent, validated on the DL3DV-10K scene dataset. It establishes feasibility on the order of a few hundred scenes and demonstrates image- and text-conditioned generation as applications. Three things remain open, and together they define this thesis. First, Can3Tok is purely geometric and photometric; it does not use per-primitive semantics, so its latent has no semantic organization. Second, it is validated at a scale of hundreds of scenes; what breaks, and what is required, when the training set is scaled to several thousand heterogeneous scenes drawn from many capture sources is not characterized. Third, an important detail of its setup, that it samples a large fixed budget of Gaussians while the splats are still being jointly optimized, differs from the common downstream situation in which the splats are pre-fitted and frozen, and the consequences of that difference for the reconstruction target are unexamined. The remainder of this thesis addresses these three gaps, with the unifying claim that they are, at bottom, questions of identifiability and structure.
